---
title: "Probit Regression Function"
author: "Elliot Hipski, Rohan, Brandon, Arnav"
date: "April 2026"
output:
  beamer_presentation: default
  ioslides_presentation:
    widescreen: true
    smaller: true
subtitle: Applications in Government & Politics
---

## Table of Contents

1. Prerequisite Knowledge
2. General Explanation
3. Mathematical Explanation
4. Implementation
5. Comparison to Other Models
6. Citations and Appendix

# Topic 1 | Prerequisite Knowledge

## Bernoulli Distribution Variance
Used for modeling binary outcomes. The variance is defined as $\text{Var}(X) = E[X^2] - (E[X])^2 = p - p^2 = p(1 - p)$ 

## General Linearized Model
A *General Linearized Model* is a model that applies a linear function to a dataset, with the general rules of mapping the input to the output y, and mapping it according to a **Link Function**.

## Gradient
A vector of partial derivatives.  It is used in mapping the log likelihood function of the linear model.  For this case, we use the weights as slopes and apply the transformation $X^T \cdot W$, where $W$ is the weights matrix. This is used in solving for the **Hessian**

## Hessian
A matrix of 2nd order partial derivatives.  For this case, however, we take the negative Expected Value of the hessian, which simplifies the matrix math.  Using the previously defined weights matrix, a formula for solving the Expected Hessian becomes (2): 
\[ X^T W X^T \]

## 

## Link Function 

# Topic 2 | General Explanation

## The Probit Mechanism

- **Distribution Assumption**: Assumes an underlying normal distribution for the latent variable.
- **Inputs**: Utilizes the CDF ($\Phi$) and PDF ($\phi$) of the standard normal distribution.
- **Predictor Matrix**: Uses a linear predictor $X\beta$, where $X$ contains a column of 1s (intercept) and a column of observations.
- **Score Calculation**: Calculated using the difference between $y$ variables and the normal PDF, scaled by the ratio of the PDF to the Bernoulli variance.
- **Iteration**: The system solves for $\beta$ by using the Hessian and Gradient to update variables until convergence (difference < tolerance).

# Topic 3 | Mathematical Explanation



# Topic 4 | Implementation {.shrink}

```{r echo=TRUE}
myprobit = function(X, y, max_iter, tolerance, verbose=FALSE){

  m = ncol(X)
  beta <- rep(0, m)
  
  for (i in 1:max_iter) {
    
    eta <- X %*% beta
    
    phat <- pnorm(eta)
    qhat <- dnorm(eta)
    
    ratio <- qhat / (phat * (1 - phat))
    score <- (y - phat) * ratio
    g <- t(X) %*% score
    
    w <- (qhat^2) / (phat * (1 - phat))
    H <- -t(X) %*% (X * as.vector(w))
    
    copy_beta = beta
    beta <- beta - solve(H, g)
    
    if (verbose){
      cat("Iteration", i, ":\n")
      j=1
      for (b in beta){
        cat("Beta", j, " = ", beta[j], "\n")
        j=j+1
      }
    }
    if ((sum(beta - copy_beta)^2) < tolerance) {
        return(beta)
        break
      }
  }
}
```


```{r echo=TRUE}
X <- model.matrix(am ~ mpg, data = mtcars)
y <- mtcars$am
myprobit(X,y,32, 0.00009, FALSE)

```

# Solving the System

## Optimization Methods
- There are 2 main methods of creating a model for probit regression:
### 1. Newton-Raphson method
We used a method similar to Newton Raphson, but rather than use the Hessian, the Expected Hessian was used, which is used for IRLS and is simpler to calculate.
### 2. Iteratively Reweighted Least Squares / Fisher Scoring
- These methods can be shown to be mathematically the same(7)
## Testing

# Topic 5 | Comparison to Other Models

## Motivation

Several methods exist for modeling binary outcomes. This section compares probit regression against three alternatives:

1. **Linear Regression (OLS)** — the standard baseline for continuous outcomes
2. **Logistic Regression** — the most widely used GLM for binary data
3. **Lasso Regression** — a penalized method that performs automatic variable selection

The goal is to understand the practical and theoretical differences between these approaches, and to identify when probit regression is the appropriate choice.

---

## Data: Donner Party Survival

We use the **Donner Party survival dataset** from the `Sleuth3` package. The binary outcome is survival status (1 = Survived, 0 = Died), with Age and Sex as predictors.

```{r comp-setup, echo=TRUE, message=FALSE, warning=FALSE}
library(Sleuth3); library(dplyr); library(ggplot2)
library(glmnet);  library(caret)

set.seed(42)

# recode response and sex
donner <- case2001 %>%
  mutate(
    mystatus = if_else(Status == "Survived", 1, 0),
    male     = if_else(Sex == "Male", 1, 0)
  )

head(donner[, c("Age", "Sex", "Status", "mystatus")], 4)
```

---

## Linear Regression on a Binary Response

OLS fits:

$$\hat{E}(Y \mid \mathbf{X}) = \hat{\beta}_0 + \hat{\beta}_1 X_1 + \cdots + \hat{\beta}_p X_p$$

When $Y \in \{0, 1\}$, this is the **Linear Probability Model (LPM)**. Since $\mathbf{X} \in \mathbb{R}$, fitted values are not constrained to $[0, 1]$, which means the model can produce predicted probabilities below 0 or above 1. This violates the basic requirement for a valid probability model and is the primary motivation for using a link function in GLMs.

---

## OLS on a Binary Response — Code

```{r lpm-fit-code, echo=TRUE, eval=FALSE}
# fit OLS
lpm_fit <- lm(mystatus ~ Age + male, data = donner)

donner <- donner %>%
  mutate(pi_lpm = predict(lpm_fit))

ggplot(donner, aes(x = Age, y = mystatus)) +
  geom_point(aes(color = Status), size = 2, alpha = 0.7) +
  geom_line(aes(y = pi_lpm, group = male,
                linetype = factor(male, labels = c("Female","Male"))),
            color = "darkred", linewidth = 1) +
  geom_hline(yintercept = c(0, 1), linetype = "dotted", color = "gray40") +
  labs(title = "Linear Probability Model: Predictions Can Leave [0, 1]",
       y = "P(Survived)", x = "Age", linetype = "Sex",
       caption = "Dotted lines mark the valid [0,1] probability bounds") +
  theme_bw()
```

---

## OLS on a Binary Response — Plot

```{r lpm-fit-plot, echo=FALSE, eval=TRUE, fig.height=5, fig.align='center'}
lpm_fit <- lm(mystatus ~ Age + male, data = donner)
donner  <- donner %>% mutate(pi_lpm = predict(lpm_fit))

ggplot(donner, aes(x = Age, y = mystatus)) +
  geom_point(aes(color = Status), size = 2, alpha = 0.7) +
  geom_line(aes(y = pi_lpm, group = male,
                linetype = factor(male, labels = c("Female","Male"))),
            color = "darkred", linewidth = 1) +
  geom_hline(yintercept = c(0, 1), linetype = "dotted", color = "gray40") +
  labs(title = "Linear Probability Model: Predictions Can Leave [0, 1]",
       y = "P(Survived)", x = "Age", linetype = "Sex",
       caption = "Dotted lines mark the valid [0,1] probability bounds") +
  theme_bw()
```

---

## OLS Predictions Outside [0, 1]

```{r lpm-damage, echo=TRUE}
lpm_preds <- predict(lpm_fit)

cat("Range of OLS predictions:", round(range(lpm_preds), 4), "\n")
cat("Predictions below 0:     ", sum(lpm_preds < 0), "\n")
cat("Predictions above 1:     ", sum(lpm_preds > 1), "\n")

# misclassification at 0.5 threshold
lpm_class <- ifelse(lpm_preds > 0.5, 1, 0)
cat("OLS misclassification rate:",
    round(mean(lpm_class != donner$mystatus), 4), "\n")
```

Even on this small dataset, OLS produces invalid probabilities. Linear regression is not appropriate for binary outcomes.

---

## Logistic Regression

Logistic regression uses the **logit link**:

$$\eta = \text{logit}(\pi) = \ln\left[\frac{\pi}{1-\pi}\right] = \mathbf{x}^T\boldsymbol{\beta} \quad \Rightarrow \quad \hat{\pi} = \frac{e^\eta}{1 + e^\eta}$$

This assumes errors follow a logistic distribution.

Probit regression uses the **inverse standard normal CDF**:

$$\eta = \Phi^{-1}(\pi) = \mathbf{x}^T\boldsymbol{\beta} \quad \Rightarrow \quad \hat{\pi} = \Phi(\mathbf{x}^T\hat{\boldsymbol{\beta}})$$

This assumes errors follow a standard normal distribution. Both are valid GLMs for binary outcomes — they differ only in their distributional assumption.

---

## Link Functions — Code

```{r link-viz-code, echo=TRUE, eval=FALSE}
eta_seq <- seq(-4, 4, length.out = 300)

df_links <- data.frame(
  eta    = rep(eta_seq, 2),
  pi     = c(plogis(eta_seq),  # inverse logit
             pnorm(eta_seq)),  # inverse probit (std normal CDF)
  Method = rep(c("Logistic (logit link)", "Probit (normal CDF)"), each = 300)
)

ggplot(df_links, aes(x = eta, y = pi, color = Method, linetype = Method)) +
  geom_line(linewidth = 1.3) +
  labs(title = "Inverse Link Functions: Logistic vs. Probit",
       x = expression(eta ~ "(Linear Predictor)"),
       y = expression(hat(pi) ~ "(Predicted Probability)")) +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  theme_bw() + theme(legend.position = "bottom")
```

---

## Link Functions — Plot

```{r link-viz-plot, echo=FALSE, eval=TRUE, fig.height=5, fig.align='center'}
eta_seq <- seq(-4, 4, length.out = 300)
df_links <- data.frame(
  eta    = rep(eta_seq, 2),
  pi     = c(plogis(eta_seq), pnorm(eta_seq)),
  Method = rep(c("Logistic (logit link)", "Probit (normal CDF)"), each = 300)
)
ggplot(df_links, aes(x = eta, y = pi, color = Method, linetype = Method)) +
  geom_line(linewidth = 1.3) +
  labs(title = "Inverse Link Functions: Logistic vs. Probit",
       x = expression(eta ~ "(Linear Predictor)"),
       y = expression(hat(pi) ~ "(Predicted Probability)")) +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  theme_bw() + theme(legend.position = "bottom")
```

The probit curve has lighter tails — it approaches 0 and 1 faster than the logistic curve.

---

## Fitting Both Models — Code

```{r fit-both-code, echo=TRUE, eval=FALSE}
# fit logistic and probit
logit_fit  <- glm(mystatus ~ Age + male, data = donner,
                  family = binomial(link = "logit"))

probit_fit <- glm(mystatus ~ Age + male, data = donner,
                  family = binomial(link = "probit"))

# compare coefficients
round(cbind(
  "OLS Coef"    = coef(lpm_fit),
  "Logit Coef"  = coef(logit_fit),
  "Probit Coef" = coef(probit_fit)
), 4)
```

---

## Fitting Both Models — Output

```{r fit-both-output, echo=FALSE, eval=TRUE}
logit_fit  <- glm(mystatus ~ Age + male, data = donner,
                  family = binomial(link = "logit"))
probit_fit <- glm(mystatus ~ Age + male, data = donner,
                  family = binomial(link = "probit"))

round(cbind(
  "OLS Coef"    = coef(lpm_fit),
  "Logit Coef"  = coef(logit_fit),
  "Probit Coef" = coef(probit_fit)
), 4)
```

The probit coefficients are smaller by a factor of roughly 1.6 to 1.8. The logistic distribution has variance $\pi^2/3 \approx 3.29$, while the standard normal has variance 1. The scales differ, but the substantive conclusions are the same.

---

## Predicted Probabilities — Code

```{r pred-compare-code, echo=TRUE, eval=FALSE}
donner <- donner %>%
  mutate(
    pi_logit  = predict(logit_fit,  type = "response"),
    pi_probit = predict(probit_fit, type = "response")
  )

ggplot(donner, aes(x = pi_logit, y = pi_probit, color = Status)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  labs(title = "Logistic vs. Probit: Predicted Probabilities (Donner Data)",
       x = "Logistic Predicted Probability",
       y = "Probit Predicted Probability") +
  theme_bw()
```

---

## Predicted Probabilities — Plot

```{r pred-compare-plot, echo=FALSE, eval=TRUE, fig.height=5, fig.align='center'}
donner <- donner %>%
  mutate(
    pi_logit  = predict(logit_fit,  type = "response"),
    pi_probit = predict(probit_fit, type = "response")
  )

ggplot(donner, aes(x = pi_logit, y = pi_probit, color = Status)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  labs(title = "Logistic vs. Probit: Predicted Probabilities (Donner Data)",
       x = "Logistic Predicted Probability",
       y = "Probit Predicted Probability") +
  theme_bw()
```

The predicted probabilities from both models are nearly identical, confirming that the choice of link function has minimal practical impact on fitted values.

---

## Model Fit Metrics

```{r fit-metrics, echo=TRUE}
cat(sprintf("%-28s %8s %8s\n", "Metric", "Logistic", "Probit"))
cat(sprintf("%-28s %8.3f %8.3f\n", "AIC",
            AIC(logit_fit), AIC(probit_fit)))
cat(sprintf("%-28s %8.3f %8.3f\n", "Residual Deviance",
            deviance(logit_fit), deviance(probit_fit)))
cat(sprintf("%-28s %8.3f %8.3f\n", "Null Deviance",
            logit_fit$null.deviance, probit_fit$null.deviance))

log_err  <- mean((donner$pi_logit  > 0.5) != donner$mystatus)
prob_err <- mean((donner$pi_probit > 0.5) != donner$mystatus)
cat(sprintf("%-28s %8.4f %8.4f\n", "Misclassification Rate",
            log_err, prob_err))
```

The two models produce nearly identical fit across every metric. The choice between logistic and probit is rarely driven by model performance.

---

## When to Prefer Probit

**1. Latent variable framework.**
Assume an unobserved variable $Y^* = \mathbf{x}^T\boldsymbol{\beta} + \varepsilon$ where $\varepsilon \sim N(0,1)$. We observe $Y = 1$ if $Y^* > 0$. This is the probit model exactly — a standard framework in economics and political science.

**2. Tail behavior.**
The normal CDF approaches 0 and 1 more rapidly than the logistic distribution. Probit is preferable when probabilities near the extremes are of scientific interest.

**3. Field convention.**
Probit is the standard choice in economics, political science, and related disciplines. For government and political science applications, probit is the conventional model.

**4. Marginal effects.**
Probit coefficients do not have an odds-ratio interpretation, but average marginal effects provide a clear and directly interpretable quantity.

---

## Marginal Effects

Probit coefficients have no direct probability interpretation on their own. The standard approach is to compute **Average Marginal Effects (AME)**:

$$\text{AME}_j = \frac{1}{n}\sum_{i=1}^{n} \phi(\mathbf{x}_i^T\hat{\boldsymbol{\beta}}) \cdot \hat{\beta}_j$$

where $\phi(\cdot)$ is the standard normal PDF.

```{r marginal-effects, echo=TRUE, message=FALSE, warning=FALSE}
library(margins)

# compute AMEs
ame <- margins(probit_fit)
summary(ame)[, c("factor", "AME", "SE", "z", "p")]
```

This gives the change in predicted probability for a one-unit change in each predictor, averaged across all observations.

---

## Lasso Regression

The logistic or probit objective function can be **penalized** to perform variable selection:

$$\hat{\boldsymbol{\beta}} = \operatorname{argmin}_{\boldsymbol{\beta}} \left\{ -\frac{2}{n}\ell(\boldsymbol{\beta}) + \lambda \sum_{j=1}^{p} |\beta_j| \right\}$$

The $\ell_1$ penalty shrinks small coefficients toward zero and eliminates irrelevant predictors entirely when $\lambda$ is large enough. This is particularly useful in high-dimensional settings such as political science data with many demographic and behavioral predictors.

---

## Lasso: Fitting the Model — Code

```{r lasso-fit-code, echo=TRUE, eval=FALSE}
# glmnet needs a matrix, not a formula
X_mat <- model.matrix(mystatus ~ Age + male, data = donner)[, -1]
y_vec <- donner$mystatus

# fit lasso over lambda path
lasso_path <- glmnet(X_mat, y_vec, family = "binomial", alpha = 1)

plot(lasso_path, xvar = "lambda", label = TRUE,
     main = "Lasso Coefficient Path (Penalized Logistic, Donner Data)")
```

---

## Lasso: Coefficient Path

```{r lasso-fit-plot, echo=FALSE, eval=TRUE, fig.height=5, fig.align='center'}
X_mat <- model.matrix(mystatus ~ Age + male, data = donner)[, -1]
y_vec <- donner$mystatus
lasso_path <- glmnet(X_mat, y_vec, family = "binomial", alpha = 1)
plot(lasso_path, xvar = "lambda", label = TRUE,
     main = "Lasso Coefficient Path (Penalized Logistic, Donner Data)")
```

---

## Lambda Selection via Cross-Validation — Code

```{r lasso-cv-code, echo=TRUE, eval=FALSE}
set.seed(42)
cv_lasso <- cv.glmnet(X_mat, y_vec,
                      family       = "binomial",
                      alpha        = 1,
                      nfolds       = 10,
                      type.measure = "class")

plot(cv_lasso, main = "10-Fold CV: Selecting Optimal Lambda")

cat("Lambda (min CV error): ", round(cv_lasso$lambda.min, 5), "\n")
cat("Lambda (1 SE rule):    ", round(cv_lasso$lambda.1se, 5), "\n")
```

---

## Lambda Selection via Cross-Validation — Plot

```{r lasso-cv-plot, echo=FALSE, eval=TRUE, fig.height=5, fig.align='center'}
set.seed(42)
cv_lasso <- cv.glmnet(X_mat, y_vec,
                      family       = "binomial",
                      alpha        = 1,
                      nfolds       = 10,
                      type.measure = "class")
plot(cv_lasso, main = "10-Fold CV: Selecting Optimal Lambda")
cat("Lambda (min CV error): ", round(cv_lasso$lambda.min, 5), "\n")
cat("Lambda (1 SE rule):    ", round(cv_lasso$lambda.1se, 5), "\n")
```

---

## Lasso Final Coefficients

```{r lasso-coefs, echo=TRUE}
cat("Lasso coefficients at lambda.min:\n")
round(as.matrix(coef(cv_lasso, s = "lambda.min")), 5)

cat("\nLasso coefficients at lambda.1se:\n")
round(as.matrix(coef(cv_lasso, s = "lambda.1se")), 5)
```

A dot (.) indicates a coefficient shrunk exactly to zero — that predictor has been removed. In a high-dimensional political dataset, lasso becomes essential for producing a parsimonious model.

---

## All Four Methods — Code

```{r full-compare-code, echo=TRUE, eval=FALSE}
ggplot(donner, aes(x = Age)) +
  geom_point(aes(y = mystatus, color = Status), alpha = 0.5, size = 1.5) +
  geom_line(aes(y = pi_lpm, group = male), color = "gray30",
            linewidth = 0.9, linetype = "dotdash") +
  geom_smooth(aes(y = mystatus), method = "glm",
              method.args = list(family = binomial(link = "logit")),
              se = FALSE, color = "steelblue", linewidth = 1) +
  geom_smooth(aes(y = mystatus), method = "glm",
              method.args = list(family = binomial(link = "probit")),
              se = FALSE, color = "firebrick", linewidth = 1, linetype = "dashed") +
  labs(title = "Predicted Probabilities: All Four Methods (Donner Data)",
       y = "P(Survived)", x = "Age",
       caption = "Blue = Logistic  |  Red = Probit  |  Gray = OLS  |  Lasso approx. Logistic") +
  theme_bw()
```

---

## All Four Methods — Plot

```{r full-compare-plot, echo=FALSE, eval=TRUE, fig.height=5, fig.align='center'}
ggplot(donner, aes(x = Age)) +
  geom_point(aes(y = mystatus, color = Status), alpha = 0.5, size = 1.5) +
  geom_line(aes(y = pi_lpm, group = male), color = "gray30",
            linewidth = 0.9, linetype = "dotdash") +
  geom_smooth(aes(y = mystatus), method = "glm",
              method.args = list(family = binomial(link = "logit")),
              se = FALSE, color = "steelblue", linewidth = 1) +
  geom_smooth(aes(y = mystatus), method = "glm",
              method.args = list(family = binomial(link = "probit")),
              se = FALSE, color = "firebrick", linewidth = 1, linetype = "dashed") +
  labs(title = "Predicted Probabilities: All Four Methods (Donner Data)",
       y = "P(Survived)", x = "Age",
       caption = "Blue = Logistic  |  Red = Probit  |  Gray = OLS  |  Lasso approx. Logistic") +
  theme_bw()
```

---

## Confusion Matrix Comparison — Code

```{r confusion-code, echo=TRUE, eval=FALSE}
threshold <- 0.5

results <- donner %>%
  mutate(
    pred_lpm    = factor(ifelse(pi_lpm    > threshold, 1, 0)),
    pred_logit  = factor(ifelse(pi_logit  > threshold, 1, 0)),
    pred_probit = factor(ifelse(pi_probit > threshold, 1, 0)),
    pred_lasso  = factor(ifelse(
      predict(cv_lasso, newx = X_mat, s = "lambda.min",
              type = "response")[, 1] > threshold, 1, 0)),
    actual = factor(mystatus)
  )

acc <- function(pred, actual) round(mean(pred == actual), 4)

cat(sprintf("%-20s %s\n", "Method",    "Accuracy"))
cat(sprintf("%-20s %s\n", "OLS (LPM)", acc(results$pred_lpm,    results$actual)))
cat(sprintf("%-20s %s\n", "Logistic",  acc(results$pred_logit,  results$actual)))
cat(sprintf("%-20s %s\n", "Probit",    acc(results$pred_probit, results$actual)))
cat(sprintf("%-20s %s\n", "Lasso",     acc(results$pred_lasso,  results$actual)))
```

---

## Confusion Matrix Comparison — Output

```{r confusion-output, echo=FALSE, eval=TRUE}
threshold <- 0.5

results <- donner %>%
  mutate(
    pred_lpm    = factor(ifelse(pi_lpm    > threshold, 1, 0)),
    pred_logit  = factor(ifelse(pi_logit  > threshold, 1, 0)),
    pred_probit = factor(ifelse(pi_probit > threshold, 1, 0)),
    pred_lasso  = factor(ifelse(
      predict(cv_lasso, newx = X_mat, s = "lambda.min",
              type = "response")[, 1] > threshold, 1, 0)),
    actual = factor(mystatus)
  )

acc <- function(pred, actual) round(mean(pred == actual), 4)

cat(sprintf("%-20s %s\n", "Method",    "Accuracy"))
cat(sprintf("%-20s %s\n", "OLS (LPM)", acc(results$pred_lpm,    results$actual)))
cat(sprintf("%-20s %s\n", "Logistic",  acc(results$pred_logit,  results$actual)))
cat(sprintf("%-20s %s\n", "Probit",    acc(results$pred_probit, results$actual)))
cat(sprintf("%-20s %s\n", "Lasso",     acc(results$pred_lasso,  results$actual)))
```

---

## Probit Confusion Matrix

```{r probit-confmat, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
confusionMatrix(
  data      = results$pred_probit,
  reference = results$actual,
  positive  = "1"
)
```

---

## Model Comparison Summary {.smaller}

| Criterion | OLS (LPM) | Logistic | **Probit** | Lasso |
|---|---|---|---|---|
| Predictions in $[0,1]$? | No | Yes | Yes | Yes |
| Link function | Identity | Logit | **$\Phi^{-1}$ (normal)** | Logit or Probit |
| Error distribution | Normal | Logistic | **Normal** | Logistic |
| Estimation method | OLS (closed form) | MLE (IRLS) | MLE (Fisher Scoring) | MLE + $\ell_1$ penalty |
| Coefficient interpretation | $\Delta\hat{y}$ | Log-odds / Odds ratio | Marginal effects (AME) | Log-odds (shrunk) |
| Variable selection? | No | No | No | **Yes** |
| Tuning required? | No | No | No | Yes ($\lambda$) |
| Preferred when... | Never | General / odds needed | Econ, politics, latent variable | Many predictors |
| Inference (p-values, CIs)? | Yes | Yes | Yes | Not standard |

---

## Key Takeaways

- **Linear regression (OLS)** is not appropriate for binary outcomes. Predicted values are not bounded to $[0,1]$ and the model is misspecified.

- **Logistic and probit regression** produce nearly identical predicted probabilities and model fit. Coefficients differ in scale by approximately 1.6x due to differing error distributions, but substantive conclusions are the same.

- **Probit is preferred** in economics and political science due to the latent variable framework and field convention. Coefficients are interpreted via average marginal effects.

- **Logistic regression is preferred** when odds ratios are the natural effect measure.

- **Lasso** extends either model with an $\ell_1$ penalty for automatic variable selection. Standard p-values and confidence intervals are not available after lasso selection.

- For government and political science applications, probit regression is the standard and appropriate choice.

# Citations | References {.shrink}

1. **Abbott, M. G.** (n.d.). *ECON 452 Note 15: Marginal Effects in Probit Models*. Queen's University. [https://econ.queensu.ca/faculty/abbott/econ452/452note15.pdf](https://econ.queensu.ca/faculty/abbott/econ452/452note15.pdf)

2. **Friedman, J., Hastie, T., & Tibshirani, R.** (2010). Regularization Paths for Generalized Linear Models via Coordinate Descent. *Journal of Statistical Software*, 33(1). [https://glmnet.stanford.edu](https://glmnet.stanford.edu)

3. **Hanck, C., Arnold, M., Gerber, A., & Schmelzer, M.** (2020). *11.2 Probit and Logit Regression*. Introduction to Econometrics with R. [https://www.econometrics-with-r.org/11.2-palr.html](https://www.econometrics-with-r.org/11.2-palr.html)

4. **Lucero, C.** (2026). *Lecture 16: Classification & Generalized Linear Models*. CMDA 4654, Virginia Tech.

5. **MetricGate.** (n.d.). *Probit vs. Logit Models for Binary Outcomes*. [https://metricgate.com/blogs/probit-vs-logit-regression/](https://metricgate.com/blogs/probit-vs-logit-regression/)

6. **RStudio.** (n.d.). *RStudio Addins*. [https://rstudio.github.io/rstudioaddins/](https://rstudio.github.io/rstudioaddins/)

7. **Taboga, M.** (2021). *Probit classification model - Maximum likelihood*. Lectures on Probability Theory and Mathematical Statistics. [https://www.statlect.com/fundamentals-of-statistics/probit-model-maximum-likelihood](https://www.statlect.com/fundamentals-of-statistics/probit-model-maximum-likelihood)

8. **Urban Institute.** (n.d.). *Logit and Probit Regression*. [https://www.urban.org/research/data-methods/data-analysis/quantitative-data-analysis/impact-analysis/regression-techniques/logit-and-probit-regression](https://www.urban.org/research/data-methods/data-analysis/quantitative-data-analysis/impact-analysis/regression-techniques/logit-and-probit-regression)

9. **Wikimedia Commons.** (2021). *File:Hokkaido.parliamentary.svg*. [https://commons.wikimedia.org/wiki/File:Hokkaido.parliamentary.svg](https://commons.wikimedia.org/wiki/File:Hokkaido.parliamentary.svg)

10. **Wikipedia contributors.** (2024). *Logistic regression*. Wikipedia, The Free Encyclopedia. [https://en.wikipedia.org/wiki/Logistic_regression](https://en.wikipedia.org/wiki/Logistic_regression)

11. **Xie, Y.** (2024). *Xaringan: Presentation Ninja* [R package]. [https://github.com/yihui/xaringan](https://github.com/yihui/xaringan)

12. **Yanai, Y.** (n.d.). *Logit and Probit Models*. Research Methods 1. [https://yukiyanai.github.io/teaching/rm1/contents/R/logit-n-probit.html](https://yukiyanai.github.io/teaching/rm1/contents/R/logit-n-probit.html)
- R Packages: `glmnet`, `caret`, `margins`, `Sleuth3`, `dplyr`, `ggplot2`

# Appendix
- Probit model - Maximum Likelihood details
- Convergence tolerance pseudocode logic
