---
title: "Probit Regression Function"
subtitle: "Applications in Government & Politics"
author: "Elliot Hipski"
date: "April 2026"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---
## Table of Contents

1. Prerequisite Knowledge
2. General Explanation
3. Mathematical Explanation
4. Implementation
5. Comparison to Other Models
6. Citations and Appendix

# Topic 1 | Prerequisite Knowledge

## Bernoulli Distribution 
Used for modeling binary outcomes. applies $\text{Var}(X) = E[X^2] - (E[X])^2 = p - p^2 = p(1 - p)$ 

## Logistic Regression
Probit is used instead of this model 

## Gradient
The vector of partial derivatives used to navigate the likelihood surface and find the maximum likelihood estimate.

## Hessian
A matrix of second-order partial derivatives that describes the local curvature of the log-likelihood function.

## Expected Hessian
Used in Fisher Scoring to simplify calculations; it is defined as being Positive Semi-Definite for the Probit model.

## Link Function 

# Topic 2 | General Explanation

## The Probit Mechanism

- **Distribution Assumption**: Assumes an underlying normal distribution for the latent variable.
- **Inputs**: Utilizes the CDF ($\Phi$) and PDF ($\phi$) of the standard normal distribution.
- **Predictor Matrix**: Uses a linear predictor $X\beta$, where $X$ contains a column of 1s (intercept) and a column of observations.
- **Score Calculation**: Calculated using the difference between $y$ variables and the normal PDF, scaled by the ratio of the PDF to the Bernoulli variance.
- **Iteration**: The system solves for $\beta$ by using the Hessian and Gradient to update variables until convergence (difference < tolerance).

# Topic 3 | Mathematical Explanation



# Topic 4 | Implementation

```{r echo=TRUE}
myprobit = function(X, y, max_iter, tolerance, verbose=FALSE){

  m = ncol(X)
  beta <- rep(0, m)
  
  for (i in 1:max_iter) {
    
    eta <- X %*% beta
    
    phat <- pnorm(eta)
    qhat <- dnorm(eta)
    
    ratio <- qhat / (phat * (1 - phat))
    score <- (y - phat) * ratio
    g <- t(X) %*% score
    
    w <- (qhat^2) / (phat * (1 - phat))
    H <- -t(X) %*% (X * as.vector(w))
    
    copy_beta = beta
    beta <- beta - solve(H, g)
    
    if (verbose){
      cat("Iteration", i, ":\n")
      j=1
      for (b in beta){
        cat("Beta", j, " = ", beta[j], "\n")
        j=j+1
      }
    }
    if ((sum(beta - copy_beta)^2) < tolerance) {
        return(beta)
        break
      }
  }
}
```


```{r echo=TRUE}
summary(mtcars)
X <- model.matrix(am ~ mpg, data = mtcars)
y <- mtcars$am
myprobit(X,y,32, 0.00009)

```

## Solving the System

### Optimization Methods

### Testing

# Topic 5 | Comparison to Other Models

## Probit vs. Logistic Regression

- **Similarities**:
  - Both are GLMs for predicting binary outcomes.
  - Both apply gradient functions and use Likelihood Ratio Tests.
  - Both are widely used in Econometrics and Political Science.
- **Differences**:
  - **Logit**: Uses the logit function ($log(p/1-p)$) as the link.
  - **Probit**: Uses the inverse normal distribution ($\Phi^{-1}$) as the link. Economists often prefer Probit.

## Linear & Lasso Regression

- **Linear Regression**:
  - *Difference*: Based on the information matrix $X^T X$ and design matrix $X$ multiplied by $y$; results interpreted via $R^2$ and ANOVA.
- **Lasso (LAR)**:
  - *Difference*: Focuses on variable selection and shrinkage; attempts to perform Least Angle Regression to find the linear combination $\beta$.

# References

## Citations
- Yanai, Y. [Logit and Probit](https://yukiyanai.github.io/teaching/rm1/contents/R/logit-n-probit.html)
- [Econometrics with R - Probit and Logit](https://www.econometrics-with-r.org/11.2-palr.html)
- Wikipedia: Logistic Regression
- Lucero: Lecture 16 (Probit Maximum Likelihood)
- R Packages: `xaringan`, `rstudioaddins`

## Appendix
- Probit model - Maximum Likelihood details
- Convergence tolerance pseudocode logic


