---
title: "ProbitTheory"
author: "Arnav Gupta"
date: "2026-04-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Probit Regression Theory

### Step 1: Linear Predictor

$$
\eta_i = x_i^\top \beta
$$

```{r, eval = FALSE}
eta <- X %*% beta
```

The linear predictor is passed through a standard normal CDF to convert it to a probability

### Step 2: Linear Predictor –\> Probability

$$
\hat{p}_i = \Phi(\eta_i) \quad \text{and} \quad \hat{q}_i = \phi(\eta_i)
$$

```{r, eval = FALSE}
phat <- pnorm(eta)
qhat <- dnorm(eta)
```

Using the linear predictor, $\hat{p}_i = \Phi(\eta_i)$ converts it into a probability between 0 and 1 for $Y_i = 1$, while $\hat{q}_i = \phi(\eta_i)$ indicates the rate of change of that probability with respect to the linear predictor at that point.

### Step 3: Score Vector

$$
s(\beta) = \sum_{i=1}^{N} \frac{\phi(x_i^\top \beta)}{\Phi(x_i^\top \beta)\left(1 - \Phi(x_i^\top \beta)\right)} \left(y_i - \Phi(x_i^\top \beta)\right) x_i
$$

The score function tells us in which direction should the coefficients $\beta$ to increase the probability of observing the data (dependent variable) given the parameters (independent variables).

The score function accomplishes that with three components.

$$
\frac{\phi(\eta_i)}{\Phi(\eta_i)\big(1 - \Phi(\eta_i)\big)}
\times (y_i - \Phi(\eta_i))
\times x_{ij}
$$

**Ratio:** tells us how sensitive the model is at that point $\eta_i = x_i^\top \beta$. At the start of the estimation, the coefficients $\beta$ are initialized to 0. At this point, the predicted probability of all events is 0.5, and the ratio takes a moderate value, meaning the model adjusts to errors and is positioned to learn.

**Residual:** compares what actually happened $y_i$ = 0 or 1 with the predicted probability. If it is positive, the model predicted too low. If it is negative, the model predicted too high.

**Gradient:** is obtained by differentiating the linear predictor with respect to the coefficients. This determines how changes in each coefficient affect the prediction. The gradient tells the model both the **direction** (increase or decrease a coefficient) and the **magnitude** of adjustment needed so that predicted probabilities better align with the observed data.

$$
\eta_i = x_i^\top \beta
$$

$$
\frac{\partial \eta_i}{\partial \beta_j} = x_{ij}
$$

```{r, eval = FALSE}
ratio <- qhat / (phat * (1 - phat))
score <- (y - phat) * ratio
g     <- t(X) %*% score
```

### Step 4: Hessian

The weights tell the model to focus on observations that are hardest to classify (those with predicted probabilities near 0.5 ex. initialization), and the Hessian uses this information to determine how large a step to take when updating the coefficients.

$$
\mathbf{H} = -\mathbf{X}^\top \mathbf{W} \mathbf{X}
$$

$$
w_i = \frac{\phi(\eta_i)^2}{\Phi(\eta_i)(1 - \Phi(\eta_i))}
$$

```{r, eval = FALSE}
w <- (qhat^2) / (phat * (1 - phat))
H <- -t(X) %*% (X * as.vector(w))
```

### Step 5: Newton-Raphson Update

$$
\theta_{new} = \theta_{old} - [\text{Hessian}]^{-1} \times [\text{Gradient}]
$$

$$
\beta^{(t+1)} = \beta^{(t)} - \mathbf{H}(\beta^{(t)})^{-1} s(\beta^{(t)})
$$

The algorithm looks at three things:

1.  Where it is now $\beta^{(t)}$
2.  Which direction improves the model? **The gradient**
3.  How aggressive the move in the direction should be? **The Hessian**

The new coefficients represent the difference between the previous coefficients and the new curvature adjusted step.

```{r, eval = FALSE}
beta <- beta - solve(H, g)
```

The computer is repeating the same update step over and over, each time improving the coefficients, until they stop changing.

## MtCars Implementation

What is the relationship between a car’s fuel efficiency (MPG) and the probability that it has a manual transmission?

```{r}
data(mtcars)

X <- model.matrix(am ~ mpg, data = mtcars)
y <- mtcars$am

beta <- c(0, 0)   # initial coefficients
```

$$
X =\begin{bmatrix}1 & 21.0 \\1 & 21.0 \\1 & 22.8 \\1 & 21.4 \\1 & 18.7 \\1 & 18.1 \\1 & 14.3 \\1 & 24.4 \\1 & 22.8 \\1 & 19.2 \\1 & 17.8 \\1 & 16.4 \\1 & 17.3 \\1 & 15.2 \\1 & 10.4 \\1 & 10.4 \\1 & 14.7 \\1 & 32.4 \\1 & 30.4 \\1 & 33.9 \\1 & 21.5 \\1 & 15.5 \\1 & 15.2 \\1 & 13.3 \\1 & 19.2 \\1 & 27.3 \\1 & 26.0 \\1 & 30.4 \\1 & 15.8 \\1 & 19.7 \\1 & 15.0 \\1 & 21.4\end{bmatrix}
$$

$$
\beta = \begin{bmatrix} \beta_0 \\ \beta_1 \end{bmatrix} = \begin{bmatrix} 0 \\ 0 \end{bmatrix}
$$

### Step 1: Linear Predictor

```{r}
eta <- X %*% beta

cat("Step 1: Linear Predictor (eta)\n")
print(head(eta))
cat("\nAll values = 0 → model has no information yet\n\n")
```

### Step 2: Linear Predictor –\> Probability

```{r}
phat <- pnorm(eta)
qhat <- dnorm(eta)

cat("Step 2:(phat)\n")
print(head(phat))

cat("\nStep 2:(qhat)\n")
print(head(qhat))

cat("\nInterpretation:\n")
cat("- phat = 0.5 → model is completely uncertain\n")
cat("- qhat ≈ 0.399 → maximum sensitivity to change\n\n")
```

### Step 3: Score Vector

```{r}
ratio <- qhat / (phat * (1 - phat))
score <- (y - phat) * ratio
g     <- t(X) %*% score

cat("Step 3: Ratio\n")
print(head(ratio))

cat("\nInterpretation: Ratio ≈ 1.596 → model is responsive to errors\n\n")

cat("Step 3: Score (first 6 observations)\n")
print(head(score))

cat("\nInterpretation:\n")
cat("- Positive → model underpredicts (needs to increase probability)\n")
cat("- Negative → model overpredicts (needs to decrease probability)\n\n")

cat("Step 3: Gradient\n")
print(g)

cat("\nInterpretation:\n")
cat("- Beta0 gradient:", g[1], "→ overall adjustment to baseline i.e. model less likely to be manual (1)\n")
cat("- Beta1 gradient:", g[2], "→ how mpg should influence prediction i.e. Positive Beta1 gradient indicates manual cars more likely to have higher MPH\n\n")
```

### Step 4: Hessian

```{r}
w <- (qhat^2) / (phat * (1 - phat))
H <- -t(X) %*% (X * as.vector(w))

cat("Step 4: Weights (first 6)\n")
print(head(w))

cat("\nInterpretation: All weights equal → every observation equally important at start\n\n")

cat("Step 4: Hessian Matrix\n")
print(H)

cat("\nInterpretation:\n")
cat("- Large magnitude values → strong curvature\n")
cat("- mpg term dominates → mpg is influential\n\n")
```

### Step 5: Newton-Raphson Update

```{r}
beta_new <- beta - solve(H, g)

cat("Step 5: Updated Beta\n")
print(beta_new)

cat("\nInterpretation:\n")
cat("- Beta0 is negative → baseline probability of manual is low\n")
cat("- Beta1 is positive → higher mpg increases probability of manual\n")
```

The computer is repeating the same update step over and over, each time improving the coefficients, until they stop changing.

#### Our Implementation of Probit Regresion:

```{r}
X <- model.matrix(am ~ mpg, data = mtcars)
y <- mtcars$am

beta <- c(0, 0)
tol  <- 1e-8

converged <- FALSE
i <- 0

while (!converged) {
  i <- i + 1
  
  eta <- X %*% beta
  phat <- pnorm(eta)
  qhat <- dnorm(eta)
  
  ratio <- qhat / (phat * (1 - phat))
  score <- (y - phat) * ratio
  g     <- t(X) %*% score
  
  w <- (qhat^2) / (phat * (1 - phat))
  H <- -t(X) %*% (X * as.vector(w))
  
  beta <- beta - solve(H, g)
  
  if (max(abs(g)) < tol) {
    converged <- TRUE
    cat("Converged at iteration", i, "\n")
    cat("Final Beta0 =", beta[1], "\n")
    cat("Final Beta1 =", beta[2], "\n")
  }
}
```

```{r}
modell <- coef(glm(am ~ mpg, data = mtcars, family = binomial(link="probit")))
modell
```

R's in-built function confirms the results from our implementation to four decimal places. The intercept (−3.91) represents the baseline latent tendency for a car to be manual when MPG is zero; it is strongly negative, meaning the model starts from a situation where automatic transmission is much more likely before considering MPG. The MPG coefficient (0.182) means that each additional unit of MPG increases the latent “manual likelihood” score, shifting the probability toward a manual transmission. Together, they imply that higher-MPG cars are increasingly likely to be manual, but the actual probability depends on how this latent score is transformed through the probit model’s normal CDF.

```{r}
model <- glm(am ~ mpg, 
             data = mtcars, 
             family = binomial(link = "probit"))

mpg_seq <- seq(min(mtcars$mpg), max(mtcars$mpg), length.out = 100)

pred_prob <- predict(model, newdata = data.frame(mpg = mpg_seq), type = "response")

plot(mtcars$mpg, mtcars$am,
     xlab = "MPG",
     ylab = "Probability of Manual (am = 1)",
     main = "Probit Regression: MPG vs Manual Transmission",
     pch = 16)

lines(mpg_seq, pred_prob, lwd = 2)
```

There is a positive relationship between MPG and the probability that a car has a manual transmission: as fuel efficiency increases, the likelihood that a car is manual increases.

The probability that a car has a manual transmission increases smoothly with MPG, with the most rapid increase occurring around 20 MPG, where the likelihood transitions from mostly automatic to mostly manual.

margins() takes the probit coefficients and converts them into marginal effects on probability, which are actually interpretable.

```{r}
library(margins)
mfx <- margins(model)
summary(mfx)
```

To answer our research question, what is the relationship between a car’s fuel efficiency (MPG) and the probability that it has a manual transmission, for each 1 MPG increase, the probability that a car is manual increases by about 4.7 percentage points (on average).

### Extension

When extending the probit model to include multiple predictors, the structure of the model remains unchanged, but the linear predictor becomes a combination of multiple variables:

$$
P(y=1) = \Phi(\beta_0 + \beta_1 \cdot \text{mpg} + \beta_2 \cdot \text{wt})
$$

Each coefficient represents a partial effect, meaning that the effect of one variable is measured while holding the other variables constant.

If we add weight to the model, the MPG coefficient captures its effect on transmission probability while holding weight constant—meaning it reflects MPG’s independent contribution rather than differences driven by weight.

### Government/Politics Dataset

Like we mentioned earlier in the presentation, probit regression is excellent for modeling situations where the result is binary. One good example is voting, where there are two possible outcomes: voted or not voted.

We considered the research question, to what extent do socioeconomic resources, specifically age, education, and income, predict voter turnout in American national elections?

We used a dataset provided by the American National Election Studies, detailing whether individuals voted or not in the 2024 election. The dataset had over 44 variables it considered, such as occupation and religion.

Translating information from the codebook, we considered the variables below.

![The variable, VCF0009z, considers if a survey accidentally talks to too many men and not enough women, the women’s answers are given a higher **weight** (they count for more than one person) and the men’s answers are given a lower **weight** (they count for less than one person). This balances the data so the final results look like a perfect "mini-version" of the real world rather than just a reflection of who happened to answer the phone.](images/clipboard-593014418.png)

We cleaned the dataset to only use the variables above

```{r}
df <- read.csv("https://drive.google.com/uc?export=download&id=12eUO5FT7h3KXgxrlCy3UwfnxdOWTcJvT", stringsAsFactors = FALSE)


vars <- c("VCF0702",  # turnout
          "VCF0101",  # age
          "VCF0110",  # education
          "VCF0114",  # income
          "VCF0301",  # party ID
          "VCF0310",  # political interest
          "VCF0009z") # weights

df_clean <- df[, vars]

df_clean[df_clean < 0] <- NA

# Dependent Variable: Turnout (1 = voted, 0 = did not vote)
df_clean$vote <- ifelse(df_clean$VCF0702 == 2, 1,
                  ifelse(df_clean$VCF0702 == 1, 0, NA))

df_clean$age <- as.numeric(df_clean$VCF0101)

df_clean$education <- factor(df_clean$VCF0110,
                            levels = c(4,3,2,1),
                            labels = c("CollegePlus", "SomeCollege", "HighSchool", "LessHS"))

df_clean$income <- as.numeric(df_clean$VCF0114)

df_clean$party_id <- factor(df_clean$VCF0301,
                           levels = c(4,3,5,2,6,1,7),
                           labels = c("Independent",
                                      "IndLeanDem",
                                      "IndLeanRep",
                                      "WeakDem",
                                      "WeakRep",
                                      "StrongDem",
                                      "StrongRep"))

df_clean$interest <- as.numeric(df_clean$VCF0310)

df_clean$weight <- df_clean$VCF0009z


df_clean <- na.omit(df_clean)
```

#### Exploratory Data Analysis

```{r}
chisq.test(table(df_clean$vote, df_clean$education))

chisq.test(table(df_clean$vote, df_clean$party_id))
```

Both chi-squared tests show that turnout is **strongly related** to education and party identification. For education, the test statistic is very large $X^2 = 3910.4$ with a p-value effectively equal to zero, meaning we can confidently reject the idea that voting behavior is independent of education level. Similarly, for party identification, the chi-squared test $X^2 = 3694.5$ also produces a near-zero p-value, indicating a strong relationship between party ID and turnout.

In simple terms, this means that people with different education levels and different political affiliations vote at different rates. Because these variables are clearly associated with turnout, they are important predictors and should be included in the model.

**Age** is included because older individuals are generally more likely to vote.

**Income** is included because individuals with higher socioeconomic status often have more resources, time, and access to information, all of which increase the likelihood of voting.

**Political interest** is a direct measure of engagement with politics, and individuals who report higher levels of interest are much more likely to participate in elections.

```{r}
model1 <- glm(vote ~ age + education + income + party_id + interest,
              data = df_clean,
              weights = weight,
              family = binomial(link = "probit"))

summary(model1)
```

```{r}
if (!require(margins)) install.packages("margins")

library(margins)

mfx <- margins(model1)
summary(mfx)
```

The results from the probit model show that socioeconomic resources—age, education, and income—are all **strong and statistically significant predictors of voter turnout** in American national elections.

First, **age** has a positive effect on turnout. The average marginal effect (AME) of 0.0039 means that each additional year of age increases the probability of voting by about **0.4 percentage points**, holding other factors constant. This suggests that older individuals are more likely to vote.

Second, **education** has a large and meaningful impact. Compared to individuals with a college degree (the baseline group), those with lower levels of education are significantly less likely to vote. For example, having only a high school education decreases the probability of voting by about **15.5 percentage points**, while having less than a high school education decreases it by about **24.5 percentage points**. This indicates that education is one of the strongest predictors of turnout.

Third, **income** also positively affects turnout. The AME of 0.0381 shows that moving up one income category increases the probability of voting by about **3.8 percentage points**, suggesting that individuals with greater financial resources are more likely to participate in elections.

**Party identification and political interest** also have strong and significant effects on voter turnout. Compared to independents, individuals who identify with a political party, especially strong partisans, are substantially more likely to vote. For example, strong Democrats and strong Republicans are about **20–22 percentage points more likely** to vote than independents. This suggests that partisan attachment increases motivation to participate in elections.

Similarly, **political interest** has one of the largest effects in the model. The average marginal effect of 0.1076 indicates that moving up one level of interest increases the probability of voting by about **10.8 percentage points**. This means that individuals who are more engaged and attentive to politics are much more likely to turn out.

Overall, while socioeconomic resources are important, these results show that political engagement and partisan identity are equally powerful predictors of turnout.
